Efficient forwards-time simulation with ancestral recombination graphs

### 2017-08-26 forwards_paper.tex
added writeup on data structures and wrote out simple WF sim

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\bibliographystyle{plainnat}

@@

\newcommand{\msprime}{\texttt{msprime}}

@@

@@

OUTLINE
\begin{itemize}
    \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
    \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
    \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
    \item review something about your efficient methods for storing/traversing tree sequence
    \item write down algorithm used in ftprime to produce msprime-ready coalescence records
    \item describe simplify algorithm
    \item comparison of speed with simupop, fwdpp
\end{itemize}

@@

\plr{two main points below: can put down neutral mutations afterwards, and could be useful to compute stats even before that}

@@

\plr{Quick review of the ARG and discussion of whether the ARG means the data structure or the reverse-time markov process (the former).}
Suppose that to the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
we add information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or \emph{ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships between any given set of samples --
only those portions of it from which those samples have actually inherited, back to their most recent common ancestors.
The assumptions of coalescent theory -- random mating and neutrality --
imply that the stochastic law of this random set forms a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder.
Although this point was in principle known since \citet{griffithsmarjoram},
only recent algorithmic advances made it possible to actually simulate the process correctly,
without approximation, across whole chromosomes \citep{msprime}.

@@

\plr{Description of what we actually need to know in the end (the trees).}

\plr{Quick review of \msprime methods: sparse trees, tree differences.}

First we recall the data storage methods used by msprime \citep[see][for more details]{msprime}.

To record the history and identity of a set of sampled genomes,
it is only necessary to record commmon ancestry (or, \emph{coalescence}) events --
most recent common ancestors of distinct segments of sampled genomes --
and the identity and location of any mutations that these most recent common ancestors carry
that are not shared by any other recorded ancestor.
\plr{(not quite right)}

The topology of the ARG
is recorded by msprime as a list of Coalescence Records, which are tuples of the form
\texttt{(left,right,parent,children)},
which indicates that the individuals listed in \texttt{children}
have inherited the segment of genome between \texttt{left} and \texttt{right}
from the ancestor \texttt{parent}.
Individuals are identified by integers,
and there is also a table recording the \emph{birth time} of each individual.

For a set of coalescence records to be consistent,
it is required that:
\begin{enumerate}
    \item The birth time of any \texttt{children} must be strictly after the birth time of the \texttt{parent}.
    \item The set of intervals on which individual \texttt{a} is a child must be disjoint, for every \texttt{a}.
    \item The set of intervals on which individual \texttt{a} is a parent must be disjoint, for every \texttt{a}.
\end{enumerate}
These ensure that the coalescence records unambiguously specify a subset of the genealogy at any location on the genome,
but do not ensure that the entire history of the samples is specified.
Furthermore, to allow for efficient algorithms, msprime also requires that
\begin{enumerate} \setcounter{enumi}{3}
    \item The samples must be numbered \texttt{0,...,n-1}, and the smallest non-sampled label must be an internal node.
        \plr{perhaps remove this one}
    \item The list of \texttt{children} in a coalescence record must be sorted.
    \item Records must be sorted in nondecreasing order by birth time of the \texttt{parent}.
    \item Birth times must be strictly greater than zero.
\end{enumerate}

\plr{Specification for tables of nodes, edgesets, sites, and mutations (migrations?).}
\plr{Calling a collection of unordered tables a ``relaxed'' tree sequence
    but will make it clear that to do anything with these must sort first.}

\begin{itemize}
    \item[Nodes:]
        (flags, population, time)
    \item[Edgesets:]
        (left,  right,  parent,  children)
    \item[Sites:]
        (position,  ancestral\_state)
    \item[Mutations:]
        (site,  node,  derived\_state)
    \item \plr{Migrations?}
\end{itemize}

\plr{Steps to record information in forwards time.}
In the forwards simulation,
we record the results of each meiosis --
for each distinctly inherited segment of offspring chromosome,
we record the parental chromosome,
the left and right endpoints,
and the identities of any new mutations that have occurred on it.
This is recorded easily, without further processing,
in the relaxed tree sequence format.

@@

\plr{Something about simplification (but not the actual algorithm)
and estimates of how much memory usage is reduced by each time it is carried out.}
The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations 
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.
\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}

@@

@@

\begin{table}
\begin{center}
    \begin{tabular}{l||r|r||r|r|}
                    & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
                    & operations  & memory & operations  & memory \\
        mate choice & $N$       &   --  &   $N$     & -- \\
        recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
    \end{tabular}
\end{center}
\end{table}

@@

Kevin

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{listings}
\bibliographystyle{plainnat}
\usepackage[T1]{fontenc}

@@

\newcommand{\msprime}{\texttt{msprime}}
\newcommand{\nodes}{\texttt{nodes}}
\newcommand{\edgesets}{\texttt{edgesets}}
\newcommand{\sites}{\texttt{sites}}
\newcommand{\mutations}{\texttt{mutations}}

@@

\emph{Note:} author order not determined

@@

\textbf{OUTLINE}
\begin{enumerate}
    \item motivate need for whole-genome fwd-time simulations; point out that we only recently have the computing power to do this
    \item explain ARG: explain that for forwards-time only need selected loci as by defn all others can be put on afterwards
    \item review something about msprime methods for storing/traversing tree sequence
    \item describe tables and write out conditions to have valid tables
    \item write down algorithm used to do simple WF simulation
    \item describe simplify algorithm
    \item back-of-the-envelope calculation to compare cost of tracking whole genomes versus putting mutations on ARG
    \item comparison of speed with simupop, fwdpp
\end{enumerate}

@@

\plr{perhaps some of this veers into methods rather than introduction}

@@

In so doing we record the \emph{population pedigree} --
the entire history of parent-offspring relationships of an entire population going back to a remote time --
as well as information encoding the genetic outcomes of each ancestral meiosis --
who inherited which parts of which parental chromosomes.
This embellished graph is known as the \emph{ancestral recombination graph}, or {ARG} \citep{griffiths}.
Combined with ancestral genotypes and the origins of new mutations,
it completely specifies the genomic sequence of any individual in the population at any time.
However, much less than the entire ARG is needed to specify relationships 
between any given set of samples --
only those portions of it from which those samples have actually inherited, 
back to their most recent common ancestors.
The assumptions of coalescent theory 
% -- random mating and neutrality --
imply that this random set describes a Markov process looked at \emph{backwards} in time,
and can hence be simulated without reference to the unnecessary remainder 
\citet{griffithsmarjoram,msprime}.
This stochastic process is also called the ``ARG'',
but we reserve this name for the embellished graph itself.
% Although this point was in principle known since \citet{griffithsmarjoram},
% only recent algorithmic advances made it possible to actually simulate the process correctly,
% without approximation, across whole chromosomes \citep{msprime}.

@@

\plr{Reminder of what we need to know in the end (the trees), and
    quick review of \msprime{} methods: sparse trees, tree differences.}

%%%%%%
\subsection*{Data structures:}

First we describe the data structures we use for recording genealogical history,
as implemented in \msprime{}.
These derive from those described in \citet{msprime},
but have been modified to remove redundancy.

To clarify terminology,
below a \emph{tree} refers to
a genealogical tree describing how a collection of individuals are related to each other.  
A \emph{tree sequence} contains information sufficient to reconstruct the genealogical
tree relating all samples to each other at any point along the genome.
In the context of a tree sequence,
\emph{nodes} refer to distinct ancestors,
and may be identified with points in the trees of a tree sequence.
Every tip or branching point in a tree is associated with a node.
A tree sequence does not necessarily carry complete genealogical information about every node,
but calling a node a \emph{sample} implies that complete information is present for that ancestor,
i.e., the node is associated with a point in every tree in the tree sequence.
Since each node represents a certain ancestor, it has a
unique ``time``, thought of as her birth time, which determines the height
of any branching points she is associated with.  A given node will be
associated with branching points of all trees across a region if that node
is the most recent common ancestor to the subtending tips across that region.  
This information is stored in the columns of a \textbf{Node Table}:
\begin{align*}
    \text{flags,} \qquad \text{population,} \qquad \text{time} \\ 
\end{align*}
where ``flags'' records other information 
(e.g., a binary mask of `1' indicates the node is a sample).
Importantly, the \textbf{node ID} of a node is given implicitly
by the (zero-based) index of its corresponding row in the Node Table.

Tree sequences are constructed by specifying over which segments of genome
which nodes inherit from which other nodes.  This information is stored by
recording the endpoints of each distinctly inherited ancestral segment,
the parental node, and a list of children nodes who have inherited that segment.
As each such record describes a collection of edges across a swatch of trees in the tree sequence,
we call these \emph{edgesets} 
and store them in the columns of an \textbf{Edgeset Table:}
\begin{align*}
    \text{left,} \qquad \text{right,} \qquad \text{parent,} \qquad \text{children}
\end{align*}

To record information about genetic variants we need to also record each mutation
and which nodes have inherited that mutation.
The tree structure takes care of inheritance -- all we need to do
is to record the highest node in the tree at the mutated site
that inherited that mutation.
As more than one mtuation may occur at a given site,
we separate this information into two tables,
first, the \textbf{Site Table} records for each variant site
\begin{align*}
    \text{position,} \qquad \text{ancestral state}
\end{align*}
Here ``position'' is a (floating point) position along the chromosome,
and ``ancestral state'' is the genotype of the root of the tree at that site.
As for nodes, \textbf{site IDs} are given implicitly
by the (zero-based) index of the rows.
Then, we record in a \textbf{Mutation Table}
\begin{align*}
    \text{site,} \qquad \text{node,} \qquad \text{derived state}
\end{align*}
in which ``site'' is the ID of the site at wihch this mutation occurred,
``node'' is the ID of the highest node that has inherited this mutation,
and ``derived state'' is the genotype at this site of any individuals inheriting this mutation,
unless another mutation occurs.
\plr{Omit migrations?}

\paragraph{Definition of valid tables}
Here are the formal requirements for a set of nodes and edgesets to make sense,
and to allow ``msprime``'s algorithms to work properly.

To disallow time travel and multiple inheritance:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
For algorithmic reasons, we also require:
\begin{enumerate} \setcounter{enumi}{2}
    \item The leftmost endpoint of each chromosome is 0.0.
    \item Node times must be strictly greater than zero.
    \item The list of offspring in an edgeset must be sorted.
    \item Edgesets must be sorted in nondecreasing time order.
    \item The set of intervals on which each individual is a parent must be disjoint.
    \item Each edgeset must contain at least two children.
\end{enumerate}
Note that since each node time is equal to the amount of time since the \emph{birth} of the
corresponding parent, time is measured in clock time, not in meioses.

A forwards-time simulation does \textbf{not} naturally emit genealogical information
satisfying requirements 5--8.
However, \msprime{} implements two algorithms that will take a set of tables
satisfying only 1--4 and produce tables satisfying all requirements.
Trivially, \texttt{sort\_tables} enforces requirements 5 and 6 and does not renumber nodes;
then, \texttt{simplify} enforces requirements 7 and 8 (and does much more; see below).

%%%%%%
\subsection*{Recording the ARG in forwards time}

To record the genealogical history of a simulation
we need to record two things for each new chromosome:
the birth time,
and the endpoints and parental IDs of each distinctly inherited segment.
This is recorded easily, without further processing,
in the tables described above.

For concreteness, here we write out in pseudocode
how to run a neutral Wright--Fisher simulation
with overlapping generations
that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L,
on which for simplicity we assume there is exactly one crossover per generation.
The probability of death per individual each generation is \texttt{death\_prob},
and the whole-chromosome mutation rate per individual is \texttt{mut\_rate}.

\paragraph{Initialize:}
We will build the tables \nodes, \edgesets, \sites, and \mutations,
and keep track of the IDs of the current generation in the list \texttt{pop}.
\begin{lstlisting}[language=Python]
for i in 0:N-1:
    nodes.add_row(time=T)
    pop[i] = i
\end{lstlisting}
(Alternatively, the tables could be initialized by the results of a coalescent simulation.)

\paragraph{Iterate:}
We then step through the generations:
\plr{Add mutations to this:}
\begin{lstlisting}[language=Python]
for t in 1:T:
    for i in 0:N-1:
        if random.uniform() > death_prob:
            new_pop[i] = pop[i]
        else:
            u = nodes.num_rows  # the ID of the new individual
            new_pop[i] = u
            nodes.add_row(time = T-t)
            a = random.sample(pop)
            b = random.sample(pop)
            bp = random.sample(0:L)
            if bp > 0:
                edgesets.add_row(left=0, right=bp,
                                  parent=a, children=(u,))
            if bp < L:
                edgesets.add_row(left=bp, right=L,
                                  parent=b, children=(u,))
    pop = new_pop
\end{lstlisting}

\paragraph{Finalize:}
to obtain a tree sequence, we need only transform the tables to the required format:
\begin{lstlisting}
sort_tables(nodes, edgesets, sites, mutations)
simplify_tables(nodes, edgesets, sites, mutations)
\end{lstlisting}
which can then be loaded into a tree sequence with \msprime.
(These two functions operate on the tables in place.)

@@

%%%%%%
\subsection*{Tree sequence simplification}

The resulting tables encode everything we need to know --
in fact, they record all of history for everyone alive at any time through the simulation.
This is much more than we need to reconstruct the genealogies and sequences
of a smaller sample of indidivuals.
Reducing this larger tree sequence to a smaller one relevant to a given set of ``sample'' nodes
we call \emph{simplification}.
Roughly, this works by tracing ancestry from the samples backwards through the recorded history,
adding node and edgeset records to the output only when coalescent events are reached.
This works exactly as in \msprime, allowing substantial re-use of algorithms;
the main difference being that parental choice, mutations, and recombination locations 
are determined by the input tree sequence
rather than randomly generated.
An example is shown in Figure \ref{fig:method_diagram}.
\plr{(trying to make the point strongly in the figure, that everything can be stored quite concisely)}

@@

%%%%%%
\subsection*{Overview of the API}

@@

% \begin{table}
% \begin{center}
%     \begin{tabular}{l||r|r||r|r|}
%                     & \multicolumn{2}{|c|}{naive} & \multicolumn{2}{|c|}{ancestry-tracking} \\
%                     & operations  & memory & operations  & memory \\
%         mate choice & $N$       &   --  &   $N$     & -- \\
%         recombination & $N\rho$ &   --  &   $N\rho$ & -- \\
%     \end{tabular}
% \end{center}
% \end{table}

@@

@@

\appendix

\section{More general method for recording the ARG}

\plr{moved from above}
